Linear instability underlying synchronous waving of marine grass in flow

### 2014-08-30 Paper/grass.tex
Included figure showing mode shape collapse
Fixed spreadsheet

\documentclass[aps,prl,twocolumn,showpacs,superscriptaddress,10pt]{revtex4-1}  % for review and submission
%\documentclass[aps,twocolumn,floatfix,prl,10pt]{revtex4-1}
%\documentclass[letterpaper,10pt,prl,twocolumn,aps]{revtex4-1}
%\documentclass[aps,preprint,floatfix,prl]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{color}
\usepackage{longtable}
\usepackage{array}
\usepackage{amssymb}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

@@

\begin{document}
\title{Linear instability underlying synchronous waving of marine grass in flow}
\author{Ravi Singh}
\affiliation{Brown University, Providence RI 02912 USA}
\author{L. Mahadevan}
\affiliation{Harvard University, Cambridge MA 02138 USA}
\author{Mahesh Bandi\footnote{work performed while visiting Brown University}}
\affiliation{Okinawa Institute of Science and Technology, Okinawa, Japan}
\author{Amala Mahadevan}
\affiliation{Woods Hold Institute of Oceanography, Woods Hole MA USA}
\author{Shreyas Mandre}
\affiliation{Brown University, Providence RI 02912 USA}

@@

@@

We use a mean field model for the coupling between the flow and the canopy. 
Drag force exerted by the grass is approximated via a continuous body force $\mathbf{f}=-N_g\mathbf{f_d}$ in the fluid momentum balance\cite{Vivoni98,Nepf99,Ghisal02,Delangre04,Delangre06} as
\begin{equation}
\rho \left(\bu_{t}+\bu.\grad\bu \right) = -\grad P+\mu\grad^{2}\bu +\mathbf{f}+\rho\mathbf{g}, \quad \grad\cdot\bu = 0,
\label{navier-stokes}
\end{equation}
where $\mathbf{f_{d}}$ is drag force per unit length of the grass blade, $N_g$ the grass number density per unit area, $\rho$ the fluid density, $\mathbf{u}$ the velocity, 
$P$ the pressure, $\mu$ the turbulent viscosity and $\mathbf{g}$ the acceleration due to gravity. 
The Reynolds number of the flow on the scale of a grass blade is expected to be O($10^2$-$10^3$), and hence the drag force is dominated by form drag, and is modeled as $\mathbf{f_{d}}=C_N \rho u_{N}^{2}d\bn + C_{T}\rho u_{T}^{2}d\bt$ where $d$ is average width of grass blade $C_{N}$ and $C_{T}$ are normal and tangential drag coefficients respectively; $\bu_{T}$, $\bu_{N}$ are velocity vector along and
normal to grass while $\bt,\bn$ being unit vector along and normal to grass. 
We expect $C_T \ll C_N$ and take $C_T=0$ for rest of analysis. 
In the field, both $C_N$ and $N_g$ vary with distance from the bottom but we do not expect these variation to be central to the mechanism and therefore take $C_N$ and $N_g$ to be constants. 
Based on previous experiments\cite{Vivoni98}, we take $C_N \approx 1$.

For consistency of analysis, we calculate the steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ as a solution of \eqref{navier-stokes}, and examine its linear stability. 
Such a steady state flow driven by constant pressure gradient satisfies $-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$, where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< 2H$.
The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=2H$) of the fluid layer, as can be seen from the comparison between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{fig:basicflow}. 
Flow consist of a region within grass of approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from balance of drag force with pressure gradient and a parabolic velocity profile in unvegetated region due to balance of viscous force and pressure gradient. 
Continuity of flow speed and shear stress at canopy top give rise to a boundary layer in the flow profile near the tip of the grass. 
The boundary layer near the canopy top results from purely local dynamics independent of the influence of the boundaries, and may therefore be identified as a free shear layer\cite{Ghisal02} in the previous explanation of monami. 
In our model, this leads to a shear layer thickness $\delta = (\Rey\Ndg)^{-1/3} H$, where $\Ndg = \left(C_N d H N_g\right)$, $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow, $U_0 = {(dP/dx)~H^2}/{\mu}$ is scale for velocity, and $\mu$ the turbulent viscosity. 
The dependence of the boundary layer thickness on the grass density $N_g$ gives us a way to systematically control it and investigate its effect on the instability mechanism.

\begin{figure}[htb]
\includegraphics[scale=1]{Grass_Base_Vivoni_shear}
\caption{Base flow profile (left )from the experiment of Vivoni\cite{Vivoni98} (Case-3A (Figure-3.6A) with 330 plants/$m^2$, 6 blades per plant, 
plant height = 16.0$\pm0.2$ and blade width of 0.28cm)
 and it's approximation with $U_0=31cm/s$ and $\delta = 0.2H$ with our model. The model of canopy is shown in the bottom and the variation of shear layer in the inset }
\label{fig:basicflow}
\end{figure}

Considering small perturbations $u, v, p$ about the base profile $U$ and $P$ respectively, equation \eqref{navier-stokes} expanded to linear order in the perturbations yields
\begin{equation}
\begin{split}
\frac{\del u}{\del t}+U\frac{\del u}{\del x}+v\frac{\del U}{\del y} &= -\frac{1}{\rho}\frac{\del p}{\del x}+\frac{\mu}{\rho}\nabla^2u-2S(y) C_{N}dN_{g}Uu, \\
\frac{\del v}{\del  t}+ U\frac{\del v}{\del x} &= -\frac{1}{\rho}\frac{\del p}{\del y}+\frac{\mu}{\rho}\nabla^2v, \hspace{0.3cm} \nabla\cdot\bu=0.
 %\nabla\cdot \bu &= 0
\end{split}
\end{equation}
These equations can be non-dimensionalized using half channel height $H$, velocity $U_0$, and the associated advection time $H/U_0$. 
We are left with three non-dimensional parameters in the model, viz. $\Rey$, $\Ndg$, and the submergence ratio of the grass $h_g/H$. 
We also use $\delta/H$ in lieu of $\Ndg$ to parametrize the grass density and help elucidate the mechanism of the instability. 
With these scaling, and using a stream function $\psi$ with $u = \psi_{y}, v= -\psi_x$ to satisfy mass balance, we seek a wave solution of the form $\left(u,v,\psi \right)= \left(\hat u, \hat v, \hat\phi \right)e^{ikx+\sigma t}$ to  obtain a modified Orr-Sommerfield equation. 
The eigenvalue $\sigma$ for a given $k$ provides growth rate and frequency associated with a perturbation of wavelength $2\pi/k$ through the solution of
% \small
\begin{equation}
\begin{split}
\left(D^2 -k^{2} \right)^2\phi &= \Rey \left[ \left({\sigma}+ikU\right) \left(D^2-k^2\right) -ikU_{yy}\right]\phi \\
&+D\left(2R \Ndg S(y) U D \phi\right),
\label{Orr-somerfield}
\end{split}
\end{equation}
where $D=d/dy$, and subject to the boundary conditions $D\phi = D^2\phi = 0$ at $y=0$ and $y=2$.

\begin{figure}[htb]
\begin{center}
\includegraphics[]{Critical_Re_vs_delta_noshear} \\
\vspace{-6mm} \hspace{-3mm}
\includegraphics[]{K_vs_shear_width_noshear}
\end{center}
\caption{Critical Reynolds number with flow for different grass height as a function of inverse shear thickness}
\label{Re_vs_delta}
\end{figure}

@@

\begin{figure*}
\begin{tabular}{cccc}
{\includegraphics[scale = 0.72]{Set4_dens28_imgsc}} &
{\includegraphics[scale = 0.72]{Set4_dens30_imgsc}} &
{\includegraphics[scale = 0.72]{Set4_dens32_imgsc}} &
{\includegraphics[scale = 0.72]{Set4_dens34_imgsc}} \\
{\includegraphics[scale = 0.72]{Set5_dens38_imgsc}} &
{\includegraphics[scale = 0.72]{Set5_dens40_imgsc}} &
{\includegraphics[scale = 0.72]{Set5_dens42_imgsc}} &
{\includegraphics[scale = 0.72]{Set5_dens46_imgsc}} \\
{\includegraphics[scale = 0.72]{Set6_dens32_imgsc}} &
{\includegraphics[scale = 0.72]{Set6_dens34_imgsc}} &
{\includegraphics[scale = 0.72]{Set6_dens36_imgsc}} &
{\includegraphics[scale = 0.72]{Set6_dens38_imgsc}}
\end{tabular}

\caption{$Re(\sigma)$ as function of wavenumber and Reynolds number for a given grass number density along with the neutral curve ($Re(\sigma)$=0), for parameters shown in the corresponding panel.  
As $\Re/\epsilon \propto N_g$ increases the unstable region splits into two labeled as ``Mode 1'' and ``Mode 2''. For $N_g$ below
a critical value, the Mode 1 sets the threshold $\Rey$ whereas above the critical $N_{g}$ onset is determined by Mode 2.}
\label{K_Re_sigma_set3}
\end{figure*}

@@

\begin{figure}[]
\includegraphics[]{Asymptotic_eig_set3.pdf}
\caption{plot of $|\phi|$ for different shear layer thickness in limit of small shear layer thickness}
\label{Asymptotic_mode}
\end{figure}

>>>>

\documentclass[aps,prl,reprint,twocolumn,showpacs,superscriptaddress,10pt]{revtex4-1}  % for review and submission
%\documentclass[aps,twocolumn,floatfix,prl,10pt]{revtex4-1}
%\documentclass[letterpaper,10pt,prl,twocolumn,aps]{revtex4-1}
%\documentclass[aps,preprint,floatfix,prl]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{color}
\usepackage{longtable}
\usepackage{array}
\usepackage{amssymb}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

@@

\begin{document}
\title{Linear instability underlying synchronous waving of marine grass in flow}
\author{Ravi Singh}
\affiliation{Brown University, Providence RI 02912 USA}
\author{L. Mahadevan}
\affiliation{Harvard University, Cambridge MA 02138 USA}
\author{Mahesh Bandi\thanks{work performed while visiting Brown University}}
\affiliation{Okinawa Institute of Science and Technology, Okinawa, Japan}
\author{Amala Mahadevan}
\affiliation{Woods Hold Institute of Oceanography, Woods Hole MA USA}
\author{Shreyas Mandre}
\affiliation{Brown University, Providence RI 02912 USA}

@@

\begin{figure}
\includegraphics[scale=1]{Grass_Base_Vivoni_shear}
\caption{Base flow profile (left )from the experiment of Vivoni\cite{Vivoni98} (Case-3A (Figure-3.6A) with 330 plants/$m^2$, 6 blades per plant, 
plant height = 16.0$\pm0.2$ and blade width of 0.28cm)
 and it's approximation with $U_0=31cm/s$ and $\delta = 0.2H$ with our model. The model of canopy is shown in the bottom and the variation of shear layer in the inset }
\label{fig:basicflow}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[]{Critical_Re_vs_delta_noshear} \\
\vspace{-6mm} \hspace{-3mm}
\includegraphics[]{K_vs_shear_width_noshear}
\end{center}
\caption{Critical Reynolds number with flow for different grass height as a function of inverse shear thickness}
\label{Re_vs_delta}
\end{figure}
We use a mean field model for the coupling between the flow and the canopy. 
Drag force exerted by the grass is approximated via a continuous body force $\mathbf{f}=-N_g\mathbf{f_d}$ in the fluid momentum balance\cite{Vivoni98,Nepf99,Ghisal02,Delangre04,Delangre06} as
\begin{equation}
\rho \left(\bu_{t}+\bu.\grad\bu \right) = -\grad P+\mu\grad^{2}\bu +\mathbf{f}+\rho\mathbf{g}, \quad \grad\cdot\bu = 0,
\label{navier-stokes}
\end{equation}
where $\mathbf{f_{d}}$ is drag force per unit length of the grass blade, $N_g$ the grass number density per unit area, $\rho$ the fluid density, $\mathbf{u}$ the velocity, 
$P$ the pressure, $\mu$ the turbulent viscosity and $\mathbf{g}$ the acceleration due to gravity. 
The Reynolds number of the flow on the scale of a grass blade is expected to be O($10^2$-$10^3$), and hence the drag force is dominated by form drag, and is modeled as $\mathbf{f_{d}}=C_N \rho u_{N}^{2}d\bn + C_{T}\rho u_{T}^{2}d\bt$ where $d$ is average width of grass blade $C_{N}$ and $C_{T}$ are normal and tangential drag coefficients respectively; $\bu_{T}$, $\bu_{N}$ are velocity vector along and
normal to grass while $\bt,\bn$ being unit vector along and normal to grass. 
We expect $C_T \ll C_N$ and take $C_T=0$ for rest of analysis. 
In the field, both $C_N$ and $N_g$ vary with distance from the bottom but we do not expect these variation to be central to the mechanism and therefore take $C_N$ and $N_g$ to be constants. 
Based on previous experiments\cite{Vivoni98}, we take $C_N \approx 1$.

For consistency of analysis, we calculate the steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ as a solution of \eqref{navier-stokes}, and examine its linear stability. 
Such a steady state flow driven by constant pressure gradient satisfies $-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$, where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< 2H$.
The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=2H$) of the fluid layer, as can be seen from the comparison between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{fig:basicflow}. 
Flow consist of a region within grass of approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from balance of drag force with pressure gradient and a parabolic velocity profile in unvegetated region due to balance of viscous force and pressure gradient. 
Continuity of flow speed and shear stress at canopy top give rise to a boundary layer in the flow profile near the tip of the grass. 
The boundary layer near the canopy top results from purely local dynamics independent of the influence of the boundaries, and may therefore be identified as a free shear layer\cite{Ghisal02} in the previous explanation of monami. 
In our model, this leads to a shear layer thickness $\delta = (\Rey\Ndg)^{-1/3} H$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow, $U_0 = {(dP/dx)~H^2}/{\mu}$ is scale for velocity, and $\mu$ the turbulent viscosity. 
The dependence of the boundary layer thickness on the grass density $N_g$ gives us a way to systematically control it and investigate its effect on the instability mechanism.

\begin{figure*}
\begin{tabular}{cccc}
{\includegraphics[scale = 0.68]{Set4_dens28_imgsc}} &
{\includegraphics[scale = 0.68]{Set4_dens30_imgsc}} &
{\includegraphics[scale = 0.68]{Set4_dens32_imgsc}} &
{\includegraphics[scale = 0.68]{Set4_dens34_imgsc}} \\
{\includegraphics[scale = 0.68]{Set5_dens38_imgsc}} &
{\includegraphics[scale = 0.68]{Set5_dens40_imgsc}} &
{\includegraphics[scale = 0.68]{Set5_dens42_imgsc}} &
{\includegraphics[scale = 0.68]{Set5_dens46_imgsc}} \\
{\includegraphics[scale = 0.68]{Set6_dens32_imgsc}} &
{\includegraphics[scale = 0.68]{Set6_dens34_imgsc}} &
{\includegraphics[scale = 0.68]{Set6_dens36_imgsc}} &
{\includegraphics[scale = 0.68]{Set6_dens38_imgsc}}
\end{tabular}
\caption{$Re(\sigma)$ as function of wavenumber and Reynolds number for a given grass number density along with the neutral curve ($Re(\sigma)$=0), for parameters shown in the corresponding panel.  
As $\Re/\epsilon \propto N_g$ increases the unstable region splits into two labeled as ``Mode 1'' and ``Mode 2''. For $N_g$ below
a critical value, the Mode 1 sets the threshold $\Rey$ whereas above the critical $N_{g}$ onset is determined by Mode 2.}
\label{K_Re_sigma_set3}
\end{figure*}

Considering small perturbations $u, v, p$ about the base profile $U$ and $P$ respectively, equation \eqref{navier-stokes} expanded to linear order in the perturbations yields
\begin{equation}
\begin{split}
% \frac{\del u}{\del t}+U\frac{\del u}{\del x}+v\frac{\del U}{\del y} &= -\frac{1}{\rho}\frac{\del p}{\del x}+\frac{\mu}{\rho}\nabla^2u-2S(y) C_{N}dN_{g}Uu, \\
% \frac{\del v}{\del  t}+ U\frac{\del v}{\del x} &= -\frac{1}{\rho}\frac{\del p}{\del y}+\frac{\mu}{\rho}\nabla^2v, \hspace{0.3cm} \nabla\cdot\bu=0.
 %\nabla\cdot \bu &= 0
\rho(u_t+U u_x+vU_y) &= -p_x+ {\mu}\nabla^2u-2S\rho C_{N}dN_{g}Uu, \\
\rho(v_t+ Uv_x) &= -p_y+ {\mu}\nabla^2v, \hspace{0.3cm} \nabla\cdot\bu=0.
 %\nabla\cdot \bu &= 0
\end{split}
\end{equation}
These equations can be non-dimensionalized using half channel height $H$, velocity $U_0$, and the associated advection time $H/U_0$. 
We are left with three non-dimensional parameters in the model, viz. $\Rey$, $\Ndg$, and the submergence ratio of the grass $h_g/H$. 
We also use $\delta/H$ in lieu of $\Ndg$ to parametrize the grass density and help elucidate the mechanism of the instability. 
With these scaling, and using a stream function $\psi$ with $u = \psi_{y}, v= -\psi_x$ to satisfy mass balance, we seek a wave solution of the form $\left(u,v,\psi \right)= \left(\hat u, \hat v, \hat\phi \right)e^{ikx+\sigma t}$ to  obtain a modified Orr-Sommerfield equation. 
The eigenvalue $\sigma$ for a given $k$ provides growth rate and frequency associated with a perturbation of wavelength $2\pi/k$ through the solution of
% \small
\begin{equation}
\begin{split}
\left(D^2 -k^{2} \right)^2\phi &= \Rey \left[ \left({\sigma}+ikU\right) \left(D^2-k^2\right) -ikU_{yy}\right]\phi \\
&+D\left(2R \Ndg S U D \phi\right),
\label{Orr-somerfield}
\end{split}
\end{equation}
where $D=d/dy$, and subject to the boundary conditions $D\phi = D^2\phi = 0$ at $y=0$ and $y=2$.

@@

@@

\begin{figure}
\includegraphics[]{Asymptotic_noshear}
\caption{plot of $|\phi|$ for different shear layer thickness in limit of small shear layer thickness}
\label{Asymptotic_mode}
\end{figure}
